---
title: "Draft: Data Documentation"
---

Here is some databases I've found that may be able to help answer this question


https://www.fisheries.noaa.gov/inport/item/46426
```{r}
######### shark data
library(readr)
Biological_Data_SBK <- read_csv("Datasets/Shark Datasets/Biological _Data_SBK.csv")
```

Who collected the data: 
The first dataset is from the Southeast Fisheries Science Center which is a part of NOAA Fisheries and National Ocean Services database.

Why was the data collected: 
The data was collected to investigate the foraging ecology of early life stages of blacktip sharks in Florida, to determine levels of resource overlap.

What is the data about:
The data is about analyzing the biological information and diet information of blacktip sharks in Florida to examine foraging ecology, bioenergetics, and trophic level. 

When was the data collected:
The data was collected from 2008 to 2010 

Where was the data collected:
The data was collected at Crooked Island Sound and Gulf of Mexico side of St. Vincent Island, Florida

How was the data generated:
The data was generated by using either a gillnet or bottom longline and bait to sample sharks and record their biological informaiton and diet information

Structure: 
.csv file

Formatting descisions: 
Day, Month, and Year each have their own separate columns

Data Validation / Quality control:
This is from a government site so it will hopefully be of good quality
some missing values

License: 
There is no Data Access Policy or Use Constraints listed

https://catalog.data.gov/dataset/shark-and-red-snapper-bottom-longline-survey
```{r}
# NOAA 
# given from Dr. VanderPlas 
library(readxl)
NMFS_BLL_data_Susan_V <- read_excel("Datasets/Shark Datasets/NMFS BLL data Susan V.xlsx")

```

Who collected the Data: 
The Southeast Fisheries Science Center which is a part of NOAA 

Why was the data collected:
Provide standardized, fisheries-independent information on the abundance and distribution of shark, snapper, and grouper species

What the data is about:
It includes details such as species identification, measurements, weights, and life history data related to age, growth, and reproduction

When the data was collected:
The data was collected from 2010 to 2024

Where the data was collected: 
The data was collected on survey sites around the coasts of the Southeast of the United States and in the Gulf of Mexico

How was the data generated:
The data was generated by using a bottom longline to collect sharks. Sampling occurred during both day and night. Live sharks were typically tagged and released unless biological samples were needed.

Structure of the Data: excel file

Formating desicions in the data:
Station Date is formatted : YYYY,MM,DD HH:MM:SS
Species name is listed as its scientific name

Data Validation / Quality Control: 
This is from a government site so it will hopefully be of good quality

License: 
No specific license is provided


https://data.ca.gov/dataset/shark-incident-database-california
```{r}
# California Department of Fish and Wildlife
library(readxl)
shark_incidents_california <- read_excel(
  "Datasets/Shark Datasets/SharkIncidents_1950_2022_220302.xlsx")

```

Who collected the Data: 
The California Department of Fish and Wildlife

Why was the data collected:
Data are added whenever a shark incident occurs, this can be more or less frequent

What the data is about:
It includes details such as information on the date, time, location, water depth, human activity, injury (if any), species of shark, and source information.

When the data was collected:
The data was collected from 1950 to 2022

Where the data was collected: 
The data was collected on the coasts of California

How was the data generated:
The data is generated whenever a shark incident occurs, this can be more or less frequent

Structure of the Data: excel file

Formating desicions in the data:
Date is formatted : YYYY-MM-DD 
Time is formatted : 0.__ hrs
Species name is listed as its common name, there are two blue sharks, one blue and one blue*

Data Validation / Quality Control: 
This is from a government site so it will hopefully be of good quality

License: 
No specific license is provided
There are no restrictions on public use

https://geo.pacioos.hawaii.edu/geoserver/web/npmv1UZSWswBsFRdH0hAuV6DPDdMGJhqufNuSr_FyK6pxar_keu6ePNLUZl5z61U7vLlwYzFYdhE0ElMNsAbjzexXAoBTRusRTKeqfkmSwA/npm5a/r_F98
```{r}
library(sf)
shark_hawaii <- st_read("Datasets/Shark Datasets/PACIOOS_WMS_ONLY-hi_pacioos_all_shark_tiger.kml")
```

Who collected the Data: 
The Pacific Islands Ocean Observing System

Why was the data collected:
Data was collected in order to track Behavioral and Oceanographic Data of tiger sharks

What the data is about:
The movements of tiger sharks 

When the data was collected:
The data was collected from 2016 to the present

Where the data was collected: 
The data was collected on the ocean coasts of Hawaii

How was the data generated:
The data is generated when a tiger shark is tagged by the data collectors

Structure of the Data: KML file

Formating desicions in the data:
Spatial KML Data, any desciption of an observation is in a spatial column that is hard to read

There is only one species recorded and that is tiger sharks 

Data Validation / Quality Control: 
This is from an educational site so it will hopefully be of good quality

License: 
No specific license is provided

These datasets contain tornado information such as latitude, longtitude, magnitude, and category : https://www.ncdc.noaa.gov/stormevents/
```{r}
######## tornado data

# NOAA / NCEI
StormEvents_details_ftp_v1_0_d2024_c20250216 <- read_csv("Datasets/Tornado_Datasets/StormEvents_details-ftp_v1.0_d2024_c20250216.csv")
(StormEvents_details_ftp_v1_0_d2024_c20250216)

StormEvents_details_ftp_v1_0_d2023_c20250216 <- read_csv("Datasets/Tornado_Datasets/StormEvents_details-ftp_v1.0_d2023_c20250216.csv")

StormEvents_details_ftp_v1_0_d2022_c20241121 <- read_csv("Datasets/Tornado_Datasets/StormEvents_details-ftp_v1.0_d2022_c20241121.csv")

StormEvents_details_ftp_v1_0_d2021_c20240716 <- read_csv("Datasets/Tornado_Datasets/StormEvents_details-ftp_v1.0_d2021_c20240716.csv")

StormEvents_details_ftp_v1_0_d2020_c20240620 <- read_csv("Datasets/Tornado_Datasets/StormEvents_details-ftp_v1.0_d2020_c20240620.csv")
```

Who collected the Data: 
The data was collected by the NCEI which is a port of NOAA

Why was the data collected:
To document tornadoes that cause harm, property damage, or economic disruption

What the data is about:
The description, location, and magnitude of tornadoes in the United States

When the data was collected:
The datasets listed here are from 1950, and from 2020-2024. 
The NCEI has been documenting tornadoes since 1950. 
Each year is contained in its own dataframe before any data cleaning is done

Where the data was collected: 
The United Sates

How was the data generated:
The National Weather Service is the main source of tornado data, which they collect through their network of weather forecast offices

Structure of the Data: 
.csv file

Formating desicions in the data:
Dates are listed as YEARMONTH: YYYYMM, and Days are listed as Start Day and End Day

Data Validation / Quality Control: 
This is from a government site so it will hopefully be of good quality
There are some missing values in the data

License: 
No specific license is provided

Here is the dataset I chose to plot coastal data: https://pubs.usgs.gov/of/2013/1284/title_page.html
```{r}
# coastal data
library(sf)
CZMP_counties_2009 <- 
  read_sf("Datasets/Coastal Datasets/CZMP_counties_2009/CZMP_counties_2009.shp")
```

Who collected the Data: 
The data was collected by the United States Geological Survey

Why was the data collected:
The data was collected to create a shapefile of 492 Coastal Zone Management Program (CZMP) counties in the United States and some territories

What the data is about:
Coastline data from the United States

When the data was collected:
The data is from coastal ground conditions in 2009

Where the data was collected: 
The United Sates

How was the data generated:
The CZMP counties shapefile was created using a three-step process involving the 2008 TIGER/Line shapefile, a NOAA list of CZMP counties, and reconciliation with historical county changes. In 2009, updates were made to align with the latest TIGER/Line data, increasing the dataset to 492 counties and 501 polygons. The shapefile was further revised following Alaska's withdrawal from the CZMP in 2011, but Alaskan county equivalents were retained for geographic reference, with a note that the data reflects 2009 conditions.

Structure of the Data: 
.shp file
There is also a lot of other files that are contained in the Coastal Datasets folder in the Datasets folder

Formating desicions in the data:
The 'geometry' column contains shape data on the latitude and longitude of the coasts that will need to be properly formatted when plotting

Data Validation / Quality Control: 
This is from a government site so it will hopefully be of good quality

License: 
No specific license is provided
From the USGS: "The USGS is committed to and is making every possible effort to ensure that all electronic and information technology developed, procured, maintained, or used by the USGS is accessible to people with disabilities, including both our employees and the customers we serve."



Suggested Data Analysis Methods : 

Use the Latitude and Longitude columns to plot the location of each shark
When the shapefile for the coastal areas are plotted, we can see where these sharks are compared to the coastal areas
Next, plot the torando start points and end points are plotted and find a way to code each tornado to a color or shape to help identify their category strength or magnitude

Finally, we can see if there is a high density of tornadoes that form close to the coast and also near shark infested waters that move inwards towards cities

